---
title: "Exercise_08_BC"
author: "Betsy Cowdery"
date: "October 27, 2014"
output: html_document
---

## Case Study: Lab 4 pine cones data

```{r}
load("data/Lab8_cone.RData")
```

###  AIC vs Likelihood Ratio Test

**Lab Report Task 1**
Compare the AICs for the two models, interpret the results for this test, and compare with the results from the LRT previously performed.
```{r, echo=FALSE}
AIC.combined = 2*lnL[1] + 2*3
AIC.treatment  = 2*(lnL[2]+lnL[3]) + 2*6
sprintf("Combined treatments ACI : %f", AIC.combined)
sprintf("Separate treatments ACI : %f", AIC.treatment)
```

#### Compare the AICs for the two models, interpret the results for this test, and compare with the results from the LRT previously performed.

### Bootstrapped Interval Estimates
```{r,echo=FALSE}
## bootstrap
nboot <- 100    		## number of bootstrap samples
npred <- 31				## number of X values you predict for
dseq  <- seq(0,30,length=npred)		## diameter sequence
mle   <- matrix(NA,nrow=nboot,ncol=3)	## storage for parameter estimates
conf.mat  <- matrix(NA,nrow=nboot,ncol=npred)	## storage for maturation
conf.cone <- matrix(NA,nrow=nboot,ncol=npred)	## storage for fecundity
pred.mat   <- matrix(NA,nrow=nboot,ncol=npred)	## storage for predictive maturation
pred.cone <- matrix(NA,nrow=nboot,ncol=npred)	## storage for predictive fecundity
```

```{r,echo=FALSE}
likfit = function(param,dia,ncone){
  a0 = param[1]
  b0 = param[2]
  b1 = param[3]
  cones = ncone > 0
  
  ## trees with cones
  dia.cone  = dia[cones > 0]                ##find just the trees with cones
  g.cone = a0 * dia.cone^2			## Fecundity fnc - g(x)
  theta.cone    = pnorm(dia.cone,b0,b1,log.p=TRUE) 	## maturation probit
  prob.cone = theta.cone + dpois(ncone[cones],g.cone,log=TRUE)
  
  ##trees with zero counts 
  dia.zero  = dia[cones == 0]
  g.zero = a0 * dia.zero^2
  theta.zero    = pnorm(dia.zero,b0,b1)   	##maturation probit
  prob.zero = log((1-theta.zero) + theta.zero*dpois(0,g.zero))
  
  return(-sum(prob.cone,prob.zero))
  }
```

**Lab Report Task 2**
Generate density and CI plots for all 3 model parameters. Include three figures and a table of parameter estimates, standard errors, and CI.

```{r, echo=FALSE}
treatment <- c("AMB","CO2")

for(n in 1:2){
  dia.t   <- b$diam[tmt==treatment[n]]   ## tree diameters
  cones.t <- (b$c00 > 0)         ## whether cones are present
  cones.t <- cones[tmt==treatment[n]]
  ncone.t <- b$c00[tmt==treatment[n]]    ## number of cones
  
  for(i in 1:nboot){
    if(i%%100 == 0) print(i)  					## progress indicator
    samp <- sample(length(dia.t),replace=T)		##Sample row indices 
    out.boot <- optim(param,likfit,method="L-BFGS-B",	## fit model to sample
                      lower=c(0.001,10,1),upper=c(1,30,30),dia=dia.t[samp],ncone=ncone.t[samp])
    mle[i,] <- out.boot$par					## store parameters
    conf.mat[i,]  <- pnorm(dseq,mle[i,2],mle[i,3])		## store model | parms
    conf.cone[i,] <- conf.mat[i,]*mle[i,1]*dseq^2
    pred.mat[i,]  <- rbinom(npred,1,conf.mat[i,])		## store pseudodata
    pred.cone[i,] <- rpois(npred,conf.cone[i,]*pred.mat[i,])
    }
  
  ints <- list(cm=conf.mat,cc=conf.cone,pm=pred.mat,pc=pred.cone) 
  assign(paste0("ints.",treatment[n]),ints)
  
  colnames(mle) = c("a0","b0","b1")
  a0.ci <- quantile(mle[,1],c(0.025,0.975))
  b0.ci <- quantile(mle[,2],c(0.025,0.975))
  b1.ci <- quantile(mle[,3],c(0.025,0.975))
  ci <- rbind(a0.ci,b0.ci,b1.ci)
  vars <- rbind(a0,b0,b1)
  # pairs(mle)
  
  ## density plots
  par(mfrow = c(1,3))
  for(j in 1:3){
    plot(density(mle[,j]),type='l',main = colnames(mle)[j])
    abline(v=ci[j,],lty=2, col = 2)
    abline(v=vars[j,n+1], col = 4)
    
    }
  
  for(v in c("mle", "conf.cone", "conf.mat", "pred.cone", "pred.mat","ci", "vars")){
    assign(paste0(v,".",treatment[n]),eval(parse(text = v)))
    }
  }

```

```{r,echo=FALSE}
a0.table <- rbind(
    c(colMeans(mle.AMB)["a0"], var(mle.AMB[,"a0"])/sqrt(length(mle.AMB[,"a0"])), ci.AMB["a0.ci",]),
    c(colMeans(mle.CO2)["a0"], var(mle.CO2[,"a0"])/sqrt(length(mle.CO2[,"a0"])), ci.CO2["a0.ci",])  )
colnames(a0.table)[2] <- "SE"
rownames(a0.table) <- c("Ambient", "Elevated")

b0.table <- rbind(
    c(colMeans(mle.AMB)["b0"], var(mle.AMB[,"b0"])/sqrt(length(mle.AMB[,"b0"])), ci.AMB["b0.ci",]),
    c(colMeans(mle.CO2)["b0"], var(mle.CO2[,"b0"])/sqrt(length(mle.CO2[,"b0"])), ci.CO2["b0.ci",])  )
colnames(b0.table)[2] <- "SE"
rownames(b0.table) <- c("Ambient", "Elevated")

b1.table <- rbind(
    c(colMeans(mle.AMB)["b1"], var(mle.AMB[,"b1"])/sqrt(length(mle.AMB[,"b1"])), ci.AMB["b1.ci",]),
    c(colMeans(mle.CO2)["b1"], var(mle.CO2[,"b1"])/sqrt(length(mle.CO2[,"b1"])), ci.CO2["b1.ci",])  )
colnames(b1.table)[2] <- "SE"
rownames(b1.table) <- c("Ambient", "Elevated")


a0.table
b0.table
b1.table
```

### Model Intervals

```{r,echo=FALSE}

ci.mat.AMB  <- apply(conf.mat.AMB,2,quantile,c(0.025,0.5,0.975))
ci.cone.AMB <- apply(conf.cone.AMB,2,quantile,c(0.025,0.5,0.975))
pi.mat.AMB  <- apply(pred.mat.AMB,2,quantile,c(0.025,0.975))
pi.cone.AMB <- apply(pred.cone.AMB,2,quantile,c(0.025,0.975))

ci.mat.CO2  <- apply(conf.mat.CO2,2,quantile,c(0.025,0.5,0.975))
ci.cone.CO2 <- apply(conf.cone.CO2,2,quantile,c(0.025,0.5,0.975))
pi.mat.CO2  <- apply(pred.mat.CO2,2,quantile,c(0.025,0.975))
pi.cone.CO2 <- apply(pred.cone.CO2,2,quantile,c(0.025,0.975)) 

```


**Lab Report Task 3**
Plot the confidence and predictive interval for the maturation process model. Include both this plot and the fecundity plot in your lab report.

```{r,echo=FALSE}
## fecundity plot
plot(b$diam,b$c00,col = b$tmt,ylim=c(0,30),main = "Fecundity Model - Ambient", xlab ="tree diameter (m)", ylab = "cone count")
lines(dseq,ci.cone.AMB[2,],col=4,lwd=3)   ## median model
lines(dseq,ci.cone.AMB[1,],col=2,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.cone.AMB[3,],col=2,lty=2,lwd=3)
lines(dseq,pi.cone.AMB[1,],col=3,lty=2,lwd=3)  ## 95% PI
lines(dseq,pi.cone.AMB[2,],col=3,lty=2,lwd=3)

cones = (b$c00 > 0) 
plot(b$diam,cones,col = b$tmt, main = "Maturation Model - Ambient", xlab ="tree diameter (m)", ylab = "presence of cones (boolean)")
lines(dseq,ci.mat.AMB[2,],col=4,lwd=3)   ## median model
lines(dseq,ci.mat.AMB[1,],col=2,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.mat.AMB[3,],col=2,lty=2,lwd=3)
lines(dseq,pi.mat.AMB[1,],col=3,lty=2,lwd=3)  ## 95% PI
lines(dseq,pi.mat.AMB[2,],col=3,lty=2,lwd=3)

## fecundity plot
plot(b$diam,b$c00,col = b$tmt,ylim=c(0,30),main = "Fecundity Model - Elevated", xlab ="tree diameter (m)", ylab = "cone count")
lines(dseq,ci.cone.CO2[2,],col=4,lwd=3)   ## median model
lines(dseq,ci.cone.CO2[1,],col=2,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.cone.CO2[3,],col=2,lty=2,lwd=3)
lines(dseq,pi.cone.CO2[1,],col=3,lty=2,lwd=3)  ## 95% PI
lines(dseq,pi.cone.CO2[2,],col=3,lty=2,lwd=3)

plot(b$diam,cones,col = b$tmt, main = "Maturation Model - Elevated", xlab ="tree diameter (m)", ylab = "presence of cones (boolean)")
lines(dseq,ci.mat.CO2[2,],col=4,lwd=3)   ## median model
lines(dseq,ci.mat.CO2[1,],col=2,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.mat.CO2[3,],col=2,lty=2,lwd=3)
lines(dseq,pi.mat.CO2[1,],col=3,lty=2,lwd=3)  ## 95% PI
lines(dseq,pi.mat.CO2[2,],col=3,lty=2,lwd=3)


```

** Lab Report Task 4 **
Repeat the bootstrap analysis for the elevated CO2 treatment. Include the figures and tables from tasks 2 & 3 for the elevated plots as well as a final plot that compares the model confidence intervals between the elevated and ambient data on one figure

```{r, echo=FALSE}

## fecundity plot
plot(b$diam,b$c00,col = "grey",ylim=c(0,30),main = "Comparison Fecundity - Elevated", xlab ="tree diameter (m)", ylab = "cone count")
lines(dseq,ci.cone.AMB[2,],col=2,lwd=3)   ## median model
lines(dseq,ci.cone.AMB[1,],col=2,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.cone.AMB[3,],col=2,lty=2,lwd=3)
lines(dseq,ci.cone.CO2[2,],col=3,lwd=3)   ## median model
lines(dseq,ci.cone.CO2[1,],col=3,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.cone.CO2[3,],col=3,lty=2,lwd=3)
legend("topleft", c("amb","elv"), lty=c(1,1), col=c(2,3))

## maturation plot
plot(b$diam,cones,col = "grey", main = "Comparison Maturation Model - Elevated", xlab ="tree diameter (m)", ylab = "presence of cones (boolean)")
lines(dseq,ci.mat.AMB[2,],col=2,lwd=3)   ## median model
lines(dseq,ci.mat.AMB[1,],col=2,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.mat.AMB[3,],col=2,lty=2,lwd=3)
lines(dseq,ci.mat.CO2[2,],col=3,lwd=3)   ## median model
lines(dseq,ci.mat.CO2[1,],col=3,lty=2,lwd=3)  ## 95% CI
lines(dseq,ci.mat.CO2[3,],col=3,lty=2,lwd=3)
legend("topleft", c("amb","elv"), lty=c(1,1), col=c(2,3))

```

**Extra Credit: Likelihood Profile Interval Estimates: Lab 3 Fire Scar data**

```{r,echo=FALSE}
load("data/Lab8_fire.RData")

c_mle = out$par[1]      ## MLE for the parameter “C”
lam_mle = out$par[2]        ## MLE for the parameter “Lambda”
contour(rp*c_mle,rp*lam_mle,z,levels=c(350,360,370,380),xlab="C",ylab="Lambda")
## add the  MLE's
abline(v=c_mle,lty=2)
abline(h=lam_mle,lty=2)

## likelihood profile
prof_lam <- function(c,lambda){   ## find the MLE holding lambda at a constant
  -sum(dweibull(firedata,c,lambda,log=TRUE))
}
prof_c <- function(lambda,c){    ## find the MLE holding C at a constant
  -sum(dweibull(firedata,c,lambda,log=TRUE))
}

## Profile for lambda

lambdas     = lam_mle*rp        ## sequence of lambda values being evaluated
lambda.like = numeric(nstep)        ## storage for lambda likelihood values
CgivenL      = numeric(nstep)     ## storage for c value that pairs with the lambda
for(i in 1:nstep){                  ## set lambda, solve for MLE
  lout <- optimize(prof_lam,interval=range(c_mle*rp),maximum=FALSE,lambda=lambdas[i])
  lambda.like[i] <- lout$objective      ## save the likelihood
  CgivenL[i] = lout$minimum ## save the MLE of c
}
```


**Lab Report Task 5**
Modify the previous section of code to calculate the vector c.like, that is found by fixing the value of C and fitting using the prof_c likelihood function.  Be aware that you will need to flip many components of the code, including the order of the two variables in the “points” command.  Turn in the following:

* R code for calculating c.like
* Likelihood surface plot that has the points for BOTH the profile of lambda and the profile of C

```{r}
cs      = c_mle*rp        ## sequence of c values being evaluated
c.like  = numeric(nstep)        ## storage for c likelihood values
LgivenC = numeric(nstep)     ## storage for the lambda value that pairs with c
for(i in 1:nstep){                  ## set c, solve for MLE
  lout <- optimize(
    prof_c,
    interval=range(lam_mle*rp),
    maximum=FALSE,
    c=cs[i]
    )
  c.like[i] <- lout$objective      ## save the likelihood
  LgivenC[i] = lout$minimum ## save the MLE of lambda
  }
contour(rp*c_mle,rp*lam_mle,z,levels=c(350,360,370,380),xlab="C",ylab="Lambda")
abline(v=c_mle,lty=2)
abline(h=lam_mle,lty=2)
lines(cs,LgivenC,type='b',pch="+",cex=0.5) ## add points to the contour plot

``````

```{r,echo=FALSE}
## Likelihood profile for each variable
c.seq = c_mle*rp
lambda.seq = lam_mle*rp
# plot(c.seq,c.like,type='l',xlab="C",ylab="Likelihood", main = "Likelihood Profile for C")
# plot(lambda.seq,lambda.like,type='l',xlab="Lambda",ylab="Likelihood", main = "Likelihood Profile for Lambda")
```

```{r,echo=FALSE}
## Find threshold deviance from the chi-square test
Dtest = qchisq(0.95,1)    ## Deviance for 95% CI
Dthresh = 2*out$value + Dtest	## Deviance threshold value
findThresh <- function(vec,Thr){	## find indices where vector crosses Thr
	which(as.logical(diff(sign(vec-Thr))))		
}

## Plot CI's
c.CI = c.seq[findThresh(c.like,Dthresh/2)]
plot(c.seq,c.like,type='l',xlab="C",ylab="Likelihood", main = "Likelihood Profile for C")
abline(h=Dthresh/2,lty=2) ## deviance is divided by 2 to convert to neg log likelihood
abline(v=c.CI, col=4)
```

```{r,echo=FALSE}

## Plot CI's
lambda.CI = lambda.seq[findThresh(lambda.like,Dthresh/2)]
plot(lambda.seq,lambda.like,type='l',xlab="Lambda",ylab="Likelihood", main = "Likelihood Profile for Lambda")
abline(h=Dthresh/2,lty=2) ## deviance is divided by 2 to convert to neg log likelihood
abline(v=lambda.CI, col=4)
```

** Lab Report Task 6**
Modify the previous section of code to calculate the CI for lambda as well. Turn in the likelihood profile plots with the CI and threshold for both variables and report the numeric values for the MLE and CI for both parameters

```{r,echo=FALSE}

CI.table <- rbind(
  c(c.CI[1],c.seq[which(c.like==min(c.like))],c.CI[2]),
  c(lambda.CI[1],lambda.seq[which(lambda.like==min(lambda.like))],lambda.CI[2])
  )

colnames(CI.table) <- c("2.5%","MLE","97.5")
rownames(CI.table) <- c("c","lambda")

CI.table
```
