---
title: "Exercise_09_BC"
author: "Betsy Cowdery"
date: "December 1, 2014"
output: html_document
---

```{r,echo=FALSE}
load("data/Lab9.RData")

## Maximum Likelihood Poisson Regression

PR1 = glm(y ~ TDR, family=poisson(link="log"))

ic=c(0,0)
LnL = function(beta){
  -sum(dpois(y,exp(beta[1] + beta[2]*TDR),log=TRUE))
}
PR2 = nlm(LnL,ic)
```
## Lab Report Task 1

**1.  Plot seedling densities as a function of TDR**
```{r,echo=FALSE}
plot(TDR,y, xlab="TDR", ylab="Seedling Counts")
```

**2.  Fit the Poisson regression model using one of the methods above and turn in  the summary output (hint: for the second method you will need to define the initial condition vector ic)**

```{r,echo=FALSE}
# Using approach 1
PR1 = glm(y ~ TDR, family=poisson(link="log"))
summary(PR1)
```

**3.  Add regression lines to the plot**
```{r,echo=FALSE}
plot(TDR,y, xlab="TDR", ylab="Seedling Counts")

PR1.fit <- sort(fitted(PR1))
PR1.idx <- sort(fitted(PR1), index.return=TRUE)[[2]]
lines(TDR[PR1.idx], PR1.fit, lwd = 3)
# or similarly
yy <- exp(coef(PR1)[1] + coef(PR1)[2]*TDR[PR1.idx])
lines(TDR[PR1.idx], yy, col=3, lty=2, lwd = 3)
```
**4.  Briefly describe how you would add model confidence and predictive intervals to these curves**

**5.  What would be an appropriate null model to compare to?  What metric would you use to compare the two models?**

**6.  Plot the calibration data of TDR vs. soil moisture.  Fit a Normal regression model to the calibration data, add the line to the plot, and report the summary table information for the fit**

```{r,echo=FALSE}
plot(TDRc,SMc, xlab="TDR", ylab="Seedling Counts")
```


```{r,echo=FALSE}

```







